Trump vs. Harris: A 2024 US Presidential Elections Modelling Forecast

### 2024-11-06 paper/paper.qmd
updated to reflect suggestions

# Install Packages if not downloaded:
required_packages <- c(
  "tidyverse", "sf", "here", "arrow", "ggplot2", "janitor", "purrr", "knitr", "kableExtra", "ggcorrplot"
)
for (p in required_packages) {
  if (!require(p, character.only = TRUE)) {
    install.packages(p, character.only = TRUE)
  }
}

# load libraries
library(tidyverse)
library(sf)
library(dplyr)
library(here)
library(arrow)
library(ggplot2)
library(janitor)
library(purrr)
library(knitr)
library(kableExtra)
library(ggcorrplot)

@@

```{r}
#| label: fig-percentage
#| fig-cap: The distribution of what percentage of support a candidate recieves in a poll
#| echo: false
#| eval: true
#| warning: false
#| message: false

# create plot for Distribution of Percentage
ggplot(cleaned_data, aes(x = Percentage)) +
  geom_histogram(
    bins = 30,
    fill = "orange",
    color = "black",
    alpha = 0.7
  ) +
  labs(
    title = "Distribution of Percentage",
    x = "Percentage",
    y = "Frequency"
  ) +
  theme_minimal()
```

```{r}
#| label: fig-candidatepercentage
#| fig-cap: Violin plot displaying the percentage of support Harris and Trump receive compared to all candidates
#| echo: false
#| eval: true
#| warning: false
#| message: false

@@

# create plot for Distribution of Percentage by candidate
ggplot(combined_data, aes(
  x = CandidateName,
  y = Percentage,
  fill = CandidateName
)) +
  geom_violin(alpha = 0.5) +
  geom_boxplot(width = 0.1, color = "black", outlier.shape = NA) +
  labs(
    title = "Percentage by Candidate",
    x = "Candidate",
    y = "Percentage"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_fill_manual(values = c("grey", "salmon", "blue")) +
  facet_wrap(~CandidateName, scales = "free_x") +
  theme_minimal()
```

@@

```{r}
#| label: fig-pollscoredistribution
#| fig-cap: Distribution of Poll Scores by Candidate with denisty overlayed
#| echo: false
#| eval: true
#| warning: false
#| message: false

@@

# create plot for Distribution of Poll Scores by candidate
ggplot(Select_candidates_cleaned_data, aes(x = PollScore)) +
  geom_histogram(aes(y = ..density..),
    bins = 30,
    fill = "blue",
    color = "black",
    alpha = 0.7
  ) +
  geom_density(color = "black", fill = "salmon", alpha = 0.6) +
  labs(
    title = "Distribution of Poll Scores by Candidate",
    x = "Poll Score",
    y = "Frequency"
  ) +
  theme_minimal() +
  facet_wrap(~CandidateName)
```

@@

```{r}
#| label: fig-samplesize
#| fig-cap: The distribution of sample sizes used in each poll
#| echo: false
#| eval: true
#| warning: false
#| message: false

# create plot for distribution of sample sizes by candidate
ggplot(cleaned_data, aes(x = SampleSize)) +
  geom_histogram(
    bins = 30,
    fill = "green",
    color = "black",
    alpha = 0.7
  ) +
  labs(
    title = "Distribution of Sample Size",
    x = "Sample Size",
    y = "Frequency"
  ) +
  theme_minimal()
```

```{r}
#| label: fig-statesample
#| fig-cap: The number of polls conducted in each state.
#| echo: false
#| eval: true
#| warning: false
#| message: false

@@

# create plot for # of polls per state
ggplot(poll_counts, aes(x = State, y = Count)) +
  geom_bar(stat = "identity") +
  labs(title = "Number of Polls by State", x = "State", y = "Number of Polls") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1, size = 6))
```

@@

```{r}
#| label: fig-pollrecency
#| fig-cap: The distribution of how recently a poll was conducted
#| echo: false
#| eval: true
#| warning: false
#| message: false

# create plot for distribution of poll recency
ggplot(cleaned_data, aes(x = PollRecency)) +
  geom_histogram(
    bins = 30,
    fill = "pink",
    color = "black",
    alpha = 0.7
  ) +
  labs(
    title = "Distribution of Poll Recency",
    x = "PollRecency",
    y = "Frequency"
  ) +
  theme_minimal()
```

```{r}
#| label: fig-numericgrade
#| fig-cap: The distribution of numeric grades assigned to each poll
#| echo: false
#| eval: true
#| warning: false
#| message: false

# create plot for distribution of numeric grades
ggplot(cleaned_data, aes(x = NumericGrade)) +
  geom_histogram(
    bins = 30,
    fill = "purple",
    color = "black",
    alpha = 0.7
  ) +
  labs(
    title = "Distribution of Numeric Grade",
    x = "Numeric Grade",
    y = "Frequency"
  ) +
  theme_minimal()
```

@@

```{r}
#| label: fig-corrmatrix
#| fig-cap: The correlation between the numerical predictor variables and the percentage of support for a candidate
#| echo: false
#| eval: true
#| warning: false
#| message: false

@@

# create correlation matrix
ggcorrplot(correlation_matrix, lab = TRUE, title = "Correlation Matrix")
```

@@

The model was implemented using R, with out-of-sample testing and Root Mean Square Error (RMSE) calculations applied to assess predictive accuracy as predicted in @tbl-rmse. By splitting the data into training and test sets, we evaluated how well the model performs on unseen data. This approach helps ensure that the model generalizes effectively to new data and avoids overfitting.

```{r}
#| label: tbl-rmse
#| tbl-cap: The RMSE for each model
#| echo: false
#| eval: true
#| warning: false
#| message: false

# read data
rmse_results <- read_parquet(here::here("data/analysis_data/rmse_results.parquet"))

# create table for rmse results per model
rmse_results %>%
  kable(
    caption = "RMSE Results for Candidate Models",
    col.names = c("Model", "Root Mean Squared Error (RMSE)"),
    digits = 2
  ) %>%
  kable_styling(latex_options = c("striped", "scale_down"))
```

@@

```{r}
#| label: tbl-popularvote
#| tbl-cap: The percentage of support Harris and Trump receive overall
#| echo: false
#| eval: true
#| warning: false
#| message: false

@@

# display avg percent and normalized percent
results %>%
  select(NationalTrumpAverage, NormalizedTrumpPercentage, NationalHarrisAverage, NormalizedHarrisPercentage) %>%
  kable(
    caption = "Predicted National Popular Vote for Donald Trump and Kamala Harris",
    col.names = c("Trump Average Predicted %", "Trump Normalized %", "Harris Average Predicted %", "Harris Normalized %"),
    digits = 2,
    booktabs = TRUE
  ) %>%
  kable_styling(latex_options = c("striped", "scale_down"))
```

```{r}
#| label: tbl-electoral
#| tbl-cap: The number of electoral college votes Harris vs Trump recieves
#| echo: false
#| eval: true
#| warning: false
#| message: false

@@

# display electoral votes
electoral_totals %>%
  kable(
    caption = "Predicted Number of Electoral Votes for Donald Trump and Kamala Harris",
    col.names = c("Trumps Total Electoral Votes", "Harris Total Electoral Votes"),
    digits = 2,
    booktabs = TRUE
  ) %>%
  kable_styling(latex_options = c("striped", "scale_down"))
```

@@

```{r}
#| label: tbl-statewinners
#| tbl-cap: The winners of each state, which determines who gets the electoral college votes for that state.
#| echo: false
#| eval: true
#| warning: false
#| message: false

@@

# display battle ground states margin in table
state_winners %>%
  select(State, TrumpPredicted, HarrisPredicted, Winner) %>%
  kable(
    caption = "Predicted Percentage of Electoral Votes Per State (Margin ≤ 4%)",
    col.names = c("State", "Trump Predicted %", "Harris Predicted %", "Winner"),
    digits = 2,
    booktabs = TRUE
  ) %>%
  kable_styling(latex_options = c("striped", "scale_down"))
```

```{r}
#| label: fig-predictedscoremap
#| fig-cap: "Mapped Model-predicted outcomes for 2024 elections by State"
#| echo: false
#| eval: true
#| warning: false
#| message: false

@@

# map electoral votes to state
map <- ggplot(merged_data, aes(x = long, y = lat, group = group)) +
  geom_polygon(aes(fill = Winner), color = "white") +
  scale_fill_manual(values = c("Donald Trump" = "red", "Kamala Harris" = "blue")) +
  labs(
    title = "2024 US Presidential Election Predictions by State",
    fill = "Predicted Winner"
  ) +
  theme_void() +
  theme(legend.position = "right")

@@

```{r}
#| label: tbl-sum
#| tbl-cap: Summary statistics for the cleaned poll data
#| echo: false
#| eval: true
#| warning: false
#| message: false

@@

# display summary stats
summary_stats %>%
  kable(
    caption = "Summary Statistics of Model Results",
    col.names = c("Average Poll Score", "Average Sample Size", "Average Percentage", "Total Polls"),
    digits = 3,
    booktabs = TRUE
  ) %>%
  kable_styling(latex_options = c("striped", "scale_down"))
```

@@

## What did we do?

@@

## What did we learn?

@@

## What are some weaknesses?

@@

```{r}
#| label: fig-swingstates
#| fig-cap: "Mapped Model-predicted outcomes for 2024 elections by Swing State"
#| echo: false
#| eval: true
#| warning: false
#| message: false

@@

# map that displays electoral college winners for swing states
map <- ggplot(merged_data, aes(x = long, y = lat, group = group)) +
  geom_polygon(aes(fill = Winner), color = "white") +
  scale_fill_manual(values = c("Donald Trump" = "red", "Kamala Harris" = "blue", "NA" = "grey")) +
  labs(
    title = "2024 US Presidential Election Battleground States",
    fill = "Predicted Winner"
  ) +
  theme_void() +
  theme(legend.position = "right")

@@

## What is left to learn? How should we proceed in the future?

@@

>>>>

# Install Packages if not downloaded:
required_packages <- c(
  "tidyverse", "sf", "here", "arrow", "ggplot2", "janitor", "purrr", "knitr", "kableExtra", "ggcorrplot", "modelsummary"
)
for (p in required_packages) {
  if (!require(p, character.only = TRUE)) {
    install.packages(p, character.only = TRUE)
  }
}

# load libraries
library(tidyverse)
library(sf)
library(dplyr)
library(here)
library(arrow)
library(ggplot2)
library(janitor)
library(purrr)
library(knitr)
library(kableExtra)
library(ggcorrplot)
library(modelsummary)

@@

```{r}
#| label: fig-percentage
#| fig-cap: This histogram shows the distribution of the percentage of support each candidate receives in various polls. The bimodal nature of the distribution suggests that candidates either receive very high or low support, with few receiving moderate levels of backing.
#| echo: false
#| eval: true
#| warning: false
#| message: false

# create plot for Distribution of Percentage
ggplot(cleaned_data, aes(x = Percentage)) +
  geom_histogram(
    bins = 30,
    fill = "orange",
    color = "black",
    alpha = 0.7
  ) +
  labs(
    x = "Percentage",
    y = "Frequency"
  ) +
  theme_minimal()
```

```{r}
#| label: fig-candidatepercentage
#| fig-cap:  Violin plot comparing the distribution of support between Donald Trump, Kamala Harris, and all other candidates. The two candidates exhibit higher-than-average support levels, with Harris and Trump leading the polls in terms of percentage of backing.
#| echo: false
#| eval: true
#| warning: false
#| message: false

@@

# create plot for Distribution of Percentage by candidate
ggplot(combined_data, aes(
  x = CandidateName,
  y = Percentage,
  fill = CandidateName
)) +
  geom_violin(alpha = 0.5) +
  geom_boxplot(width = 0.1, color = "black", outlier.shape = NA) +
  labs(
    x = "Candidate",
    y = "Percentage"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_fill_manual(values = c("grey", "salmon", "blue")) +
  facet_wrap(~CandidateName, scales = "free_x") +
  theme_minimal()
```

@@

```{r}
#| label: fig-pollscoredistribution
#| fig-cap: This figure shows the distribution of poll scores, which reflect the credibility of pollsters. A higher concentration of negative scores suggests that more reputable pollsters dominate the dataset, with credibility generally improving as the score becomes more negative.
#| echo: false
#| eval: true
#| warning: false
#| message: false

@@

# create plot for Distribution of Poll Scores by candidate
ggplot(Select_candidates_cleaned_data, aes(x = PollScore)) +
  geom_histogram(aes(y = ..density..),
    bins = 30,
    fill = "blue",
    color = "black",
    alpha = 0.7
  ) +
  geom_density(color = "black", fill = "salmon", alpha = 0.6) +
  labs(
    x = "Poll Score",
    y = "Frequency"
  ) +
  theme_minimal() +
  facet_wrap(~CandidateName)
```

@@

```{r}
#| label: fig-samplesize
#| fig-cap: This histogram illustrates the distribution of sample sizes across all polls. The majority of the polls have sample sizes under 2,500.
#| echo: false
#| eval: true
#| warning: false
#| message: false

# create plot for distribution of sample sizes by candidate
ggplot(cleaned_data, aes(x = SampleSize)) +
  geom_histogram(
    bins = 30,
    fill = "green",
    color = "black",
    alpha = 0.7
  ) +
  labs(
    x = "Sample Size",
    y = "Frequency"
  ) +
  theme_minimal()
```

```{r}
#| label: fig-statesample
#| fig-cap: This bar chart shows the number of polls conducted across U.S. states. Notably, battleground states have a higher number of polls conducted, reflecting their significance in the election.
#| echo: false
#| eval: true
#| warning: false
#| message: false

@@

# create plot for # of polls per state
ggplot(poll_counts, aes(x = State, y = Count)) +
  geom_bar(stat = "identity") +
  labs(x = "State", y = "Number of Polls") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1, size = 6))
```

@@

```{r}
#| label: fig-pollrecency
#| fig-cap: This histogram shows the distribution of poll recency, indicating that most polls were conducted relatively recently, with fewer polls extending back to the 2020 election cycle.
#| echo: false
#| eval: true
#| warning: false
#| message: false

# create plot for distribution of poll recency
ggplot(cleaned_data, aes(x = PollRecency)) +
  geom_histogram(
    bins = 30,
    fill = "pink",
    color = "black",
    alpha = 0.7
  ) +
  labs(
    x = "PollRecency",
    y = "Frequency"
  ) +
  theme_minimal()
```

```{r}
#| label: fig-numericgrade
#| fig-cap: This histogram displays the distribution of numeric grades assigned to each poll, with higher numeric grades reflecting more reliable and accurate polling sources.
#| echo: false
#| eval: true
#| warning: false
#| message: false

# create plot for distribution of numeric grades
ggplot(cleaned_data, aes(x = NumericGrade)) +
  geom_histogram(
    bins = 30,
    fill = "purple",
    color = "black",
    alpha = 0.7
  ) +
  labs(
    x = "Numeric Grade",
    y = "Frequency"
  ) +
  theme_minimal()
```

@@

```{r}
#| label: fig-corrmatrix
#| fig-cap: This correlation matrix visualizes the relationships between key numerical predictor variables and the outcome variable, Percentage. Notable correlations include a strong negative relationship between PollScore and NumericGrade, and a weak positive correlation between PollRecency and Percentage.
#| echo: false
#| eval: true
#| warning: false
#| message: false

@@

# create correlation matrix
ggcorrplot(correlation_matrix, lab = TRUE)
```

@@

The model was implemented using R, with out-of-sample testing and Root Mean Square Error (RMSE) calculations applied to assess predictive accuracy as predicted in @tbl-keyestimates. By splitting the data into training and test sets, we evaluated how well the model performs on unseen data. This approach helps ensure that the model generalizes effectively to new data and avoids overfitting.

For **PollScore**, the Trump model has a coefficient of -0.745, while the Harris model shows a coefficient of 0.781, indicating an inverse relationship for Trump and a positive relationship for Harris with the outcome. The standard errors for these coefficients are 0.266 and 0.350, respectively.

For **PollRecency**, the coefficients are -0.003 for Trump and -0.006 for Harris, with both models showing very small negative relationships. The standard errors are minimal, suggesting these estimates are fairly precise.

The **NumericGrade** variablity has a negative coefficient of -0.124 for Trump and a positive coefficient of 1.458 for Harris. This suggests a different relationship between the numeric grade and the outcome for each model, with Harris showing a stronger positive effect.

Other model summary statistics include a sample size of 2117 for the Trump model and 2255 for the Harris model. The R² values indicate that the Trump model explains 64.5% of the variance, while the Harris model explains 51.5%, suggesting the Trump model provides a better fit. The Adjusted R² values are slightly lower but follow the same trend.

The AIC, BIC, Log-Likelihood, and RMSE values all indicate that the Trump model is preferred over the Harris model in terms of model fit, as the Trump model has lower values for these selected metrics.

The full summary statistics for the Harris and Trump models are displayed in @tbl-modelestimatefull in @sec-appx.

```{r}
#| label: tbl-keyestimates
#| tbl-cap: Summary of key model estimates for Donald Trump and Kamala Harris, including coefficients for predictors like Poll Score, Poll Recency, and Numeric Grade, with standard errors for each estimate. Model performance statistics, such as sample size, R², and adjusted R², are also displayed.
#| echo: false
#| eval: true
#| warning: false
#| message: false

# load data and model
lm_model_trump <- readRDS(here::here("models/lm_model_trump.rds"))
lm_model_harris <- readRDS(here::here("models/lm_model_harris.rds"))

model_list <- list("Trump Model" = lm_model_trump, "Harris Model" = lm_model_harris)
modelsummary(model_list, 
             output = "kableExtra", 
             statistic = "std.error", 
             coef_map = c("PollScore" = "Poll Score", 
                          "PollRecency" = "Poll Recency", 
                          "NumericGrade" = "Numeric Grade",
                          "SampleSize" = "Sample Size")) %>%
  kable_styling(latex_options = c("striped", "scale_down"))
```

@@

```{r}
#| label: tbl-popularvote
#| tbl-cap: Predicted average and normalized national popular vote percentages for Donald Trump and Kamala Harris based on model outcomes.
#| echo: false
#| eval: true
#| warning: false
#| message: false

@@

# display avg percent and normalized percent
results %>%
  select(NationalTrumpAverage, NormalizedTrumpPercentage, NationalHarrisAverage, NormalizedHarrisPercentage) %>%
  kable(
    col.names = c("Trump Average Predicted %", "Trump Normalized %", "Harris Average Predicted %", "Harris Normalized %"),
    digits = 2,
    booktabs = TRUE
  ) %>%
  kable_styling(latex_options = c("striped", "scale_down"))
```

```{r}
#| label: tbl-electoral
#| tbl-cap:  Projected electoral vote totals for Donald Trump and Kamala Harris, based on state-by-state predictions derived from the model. The table summarizes the total electoral votes each candidate is expected to win.
#| echo: false
#| eval: true
#| warning: false
#| message: false

@@

# display electoral votes
electoral_totals %>%
  kable(
    col.names = c("Trumps Total Electoral Votes", "Harris Total Electoral Votes"),
    digits = 2,
    booktabs = TRUE
  ) %>%
  kable_styling(latex_options = c("striped", "scale_down"))
```

@@

```{r}
#| label: tbl-statewinners
#| tbl-cap: Summary of predicted electoral outcomes in battleground states where the margin of predicted support between Donald Trump and Kamala Harris is 4% or less, showing the predicted percentage of the vote for each candidate and the winner in each state.
#| echo: false
#| eval: true
#| warning: false
#| message: false

@@

# display battle ground states margin in table
state_winners %>%
  select(State, TrumpPredicted, HarrisPredicted, Winner) %>%
  kable(
    col.names = c("State", "Trump Predicted %", "Harris Predicted %", "Winner"),
    digits = 2,
    booktabs = TRUE
  ) %>%
  kable_styling(latex_options = c("striped", "scale_down"))
```

```{r}
#| label: fig-predictedscoremap
#| fig-cap: A map illustrating the predicted outcomes of the 2024 U.S. presidential election by state, based on the model's projections. States are color-coded by the predicted winner, with red representing Donald Trump and blue representing Kamala Harris.
#| echo: false
#| eval: true
#| warning: false
#| message: false

@@

# map electoral votes to state
map <- ggplot(merged_data, aes(x = long, y = lat, group = group)) +
  geom_polygon(aes(fill = Winner), color = "white") +
  scale_fill_manual(values = c("Donald Trump" = "red", "Kamala Harris" = "blue")) +
  labs(
    fill = "Predicted Winner"
  ) +
  theme_void() +
  theme(legend.position = "right")

@@

```{r}
#| label: tbl-sum
#| tbl-cap: Summary statistics of the cleaned polling data, including an average poll score of -0.379, an average sample size of 1,605, and 33.68% average support across 15,829 polls.
#| echo: false
#| eval: true
#| warning: false
#| message: false

@@

# display summary stats
summary_stats %>%
  kable(
    col.names = c("Average Poll Score", "Average Sample Size", "Average Percentage", "Total Polls"),
    digits = 3,
    booktabs = TRUE
  ) %>%
  kable_styling(latex_options = c("striped", "scale_down"))
```

@@

## Predicting 2024 U.S. Election Outcomes Using State-by-State Polling Data

@@

## Regional Polarization and the Crucial Role of Battleground States in Election Outcomes

@@

## Limitations of the Model: Data Bias, Shifting Voter Demographics, and Excluded Third-Party Candidates

@@

```{r}
#| label: fig-swingstates
#| fig-cap: A map illustrating the predicted outcomes of the 2024 U.S. presidential election in battleground states where the margin of predicted support between Donald Trump and Kamala Harris is 4% or less.
#| echo: false
#| eval: true
#| warning: false
#| message: false

@@

# map that displays electoral college winners for swing states
map <- ggplot(merged_data, aes(x = long, y = lat, group = group)) +
  geom_polygon(aes(fill = Winner), color = "white") +
  scale_fill_manual(values = c("Donald Trump" = "red", "Kamala Harris" = "blue", "NA" = "grey")) +
  labs(
    fill = "Predicted Winner"
  ) +
  theme_void() +
  theme(legend.position = "right")

@@

## Next Steps: Enhance Prediction Accuracy with Real-Time Data, Demographic Changes, and Third-Party Voting Trend**s**

@@

## Additional Tables & Figures

```{r}
#| label: tbl-modelestimatefull
#| tbl-cap: Full model estimates for the Trump and Harris models, including standard errors for each model parameter.
#| echo: false
#| eval: true
#| warning: false
#| message: false
model_list <- list("Trump Model" = lm_model_trump, "Harris Model" = lm_model_harris)
modelsummary(model_list, 
             output = "kableExtra", 
             statistic = "std.error",
             longtable = TRUE) %>%
  kable_styling(latex_options = c("striped", "scale_down"))
```
